---
title: "Project 8 Template"
output: pdf_document
---

```{r}
# Add to this package list for additional SL algorithms
pacman::p_load(
  tidyverse,
  ggthemes,
  ltmle,
  tmle,
  SuperLearner,
  tidymodels,
  caret,
  dagitty,
  ggdag,
  here)

heart_disease <- read_csv(here('heart_disease_tmle.csv'))
```

# Introduction

Heart disease is the leading cause of death in the United States, and treating it properly is an important public health goal. However, it is a complex disease with several different risk factors and potential treatments. Physicians typically recommend changes in diet, increased exercise, and/or medication to treat symptoms, but it is difficult to determine how effective any one of these factors is in treating the disease. In this project, you will explore SuperLearner, Targeted Maximum Likelihood Estimation (TMLE), and Longitudinal Targeted Maximum Likelihood Estimation (LTMLE). Using a simulated dataset, you will explore whether taking blood pressure medication reduces mortality risk. 

# Data

This dataset was simulated using R (so it does not come from a previous study or other data source). It contains several variables:

\begin{itemize}
    \item \textbf{blood\_pressure\_medication}: Treatment indicator for whether the individual took blood pressure medication (0 for control, 1 for treatment)
    \item \textbf{mortality}: Outcome indicator for whether the individual passed away from complications of heart disease (0 for no, 1 for yes)
    \item \textbf{age}: Age at time 1
    \item \textbf{sex\_at\_birth}: Sex assigned at birth (0 female, 1 male)
    \item \textbf{simplified\_race}: Simplified racial category. (1: White/Caucasian, 2: Black/African American, 3: Latinx, 4: Asian American, \newline 5: Mixed Race/Other)
    \item \textbf{income\_thousands}: Household income in thousands of dollars
    \item \textbf{college\_educ}: Indicator for college education (0 for no, 1 for yes)
    \item \textbf{bmi}: Body mass index (BMI)
    \item \textbf{chol}: Cholesterol level
    \item \textbf{blood\_pressure}: Systolic blood pressure 
    \item \textbf{bmi\_2}: BMI measured at time 2
    \item \textbf{chol\_2}: Cholesterol measured at time 2
    \item \textbf{blood\_pressure\_2}: BP measured at time 2
    \item \textbf{blood\_pressure\_medication\_2}: Whether the person took treatment at time period 2 
\end{itemize}

For the "SuperLearner" and "TMLE" portions, you can ignore any variable that ends in "\_2", we will reintroduce these for LTMLE.

# SuperLearner

## Modeling

Fit a SuperLearner model to estimate the probability of someone dying from complications of heart disease, conditional on treatment and the relevant covariates. Do the following:

\begin{enumerate}
    \item Choose a library of at least 5 machine learning algorithms to evaluate. \textbf{Note}: We did not cover how to hyperparameter tune constituent algorithms within SuperLearner in lab, but you are free to do so if you like (though not required to for this exercise). 
    \item Split your data into train and test sets.
    \item Train SuperLearner
    \item Report the risk and coefficient associated with each model, and the performance of the discrete winner and SuperLearner ensemble
    \item Create a confusion matrix and report your overall accuracy, recall, and precision
\end{enumerate}

```{r}
# Fit SuperLearner Model
listWrappers()

## sl lib
SL.libraries = c("SL.glm", "SL.mean", "SL.glm.interaction", "SL.nnet", "SL.svm")

## Train/Test split
# initial_split function from tidymodels/rsample
heart_split <- initial_split(heart_disease, prop = 3/4)

# Declare the training set with rsample::training()
train <- training(heart_split)

# y_train is mortality
y_train <- train$mortality

# x_train is everything but the outcome  
x_train <- train %>%
  select(-mortality & -ends_with("_2"))

# Do the same procedure with the test set
test <- testing(heart_split)

y_test <- test$mortality

x_test <- test %>%
  select(-mortality & -ends_with("_2"))

## Train SuperLearner
sl <- SuperLearner(Y = y_train,
                   X = x_train,
                   family = binomial(),
                   SL.library = SL.libraries)

## Risk and Coefficient of each model
sl

## Discrete winner and superlearner ensemble performance
preds <- predict(sl,
                 x_test,
                 onlySL = TRUE)

# start with y_test
validation <- y_test %>%
  # add our predictions
  bind_cols(preds$pred[,1]) %>%
  # rename columns
  rename(obs = `...1`,
         pred = `...2`) %>%
  mutate(pred = ifelse(pred >= .5, 
                           1,
                           0))

## Confusion Matrix
caret::confusionMatrix(as.factor(validation$pred),
                       as.factor(validation$obs))
```

## Discussion Questions

\begin{enumerate}
    \item Why should we, in general, prefer the SuperLearner ensemble to the discrete winner in cross-validation? Or in other words, what is the advantage of "blending" algorithms together and giving them each weights, rather than just using the single best algorithm (with best being defined as minimizing risk)?
    The primary benefits hark back to the bias-variance tradeoff that we learned about last semester and can ultimately result in better predictions. The SuperLearner ensemble can make better predictions and achieve better performance than any single contributing model (including the best performing one). Ensemble methods reduce the variance of the prediction error by adding bias.
\end{enumerate}

# Targeted Maximum Likelihood Estimation

## Causal Diagram

TMLE requires estimating two models:

\begin{enumerate}
    \item The outcome model, or the relationship between the outcome and the treatment/predictors, $P(Y|(A,W)$.
    \item The propensity score model, or the relationship between assignment to treatment and predictors $P(A|W)$
\end{enumerate}

Using ggdag and daggity, draw a directed acylcic graph (DAG) that describes the relationships between the outcome, treatment, and covariates/predictors. Note, if you think there are covariates that are not related to other variables in the dataset, note this by either including them as freestanding nodes or by omitting them and noting omissions in your discussion.

```{r}
# DAG for TMLE
theme_set(theme_dag())
source("pretty_dag.R")

# dag with separate demographic factors
# dag <- dagify(Y ~ A,
  #     Y ~ W1,
  #     A ~ W1,
  #     A ~ W2,
  #     Y ~ W2,
  #     A ~ W3,
  #     Y ~ W3,
  #     A ~ W4,
  #     Y ~ W4,
  #     Y ~ U,
  #     A ~ U,
  #     Y ~ U_W1,
  #     A ~ U_W1,
  #     Y ~ U_W2,
  #     A ~ U_W2,
  #     Y ~ U_W3,
  #     A ~ U_W3,
  #     W1 ~ U_W1,
  #     W2 ~ U_W2,
  #     W3 ~ U_W3,
  #     W4 ~ U,
  #    exposure = "A",
  #     outcome = "Y",
  #     labels = c("Y" = "mortality risk", "A" = "blood pressure \n medication", "W1" = "age", "W2" = "race", "W3" = "income", "W4" = "health covariates \n (e.g., blood pressure, cholesterol)", "U" = "background health \n exogenous factors"),
   #           coords = list(x = c(A = 0, Y = 3, W1 = 1, U = 2.5, W2 = 1.5, W3 = 2, W4 = 2.5, U_W1 = 0.5, U_W2 = 1, U_W3 = 1.5),
    #                        y = c(A = 1, Y = 1, W1 = 2, U = 3, W2 = 2, W3 = 2, W4 = 2, U_W1 = 3, U_W2 = 3, U_W3 = 3)))

# ggdag(dag, use_labels = "label") +
#  theme_dag()

# dag with combined demographic factors
dag <- dagify(Y ~ A,
       Y ~ W1,
       A ~ W1,
       A ~ W2,
       Y ~ W2,
       W2 ~ W1,
       Y ~ U,
       A ~ U,
       Y ~ U_W1,
       A ~ U_W1,
       Y ~ U_W2,
       A ~ U_W2,
       W1 ~ U_W1,
       W2 ~ U_W2,
       exposure = "A",
       outcome = "Y",
       labels = c("Y" = "mortality risk", "A" = "blood pressure \n medication", "W1" = "demographic covariates", "W2" = "health covariates \n (e.g., blood pressure, cholesterol)", "U_W1" = "background demographic \n exogenous factors", "U_W2" = "background health \n exogenous factors"),
              coords = list(x = c(A = 0, Y = 3, W1 = 1, W2 = 2, U_W1 = 0.5, U_W2 = 2.5),
                            y = c(A = 1, Y = 1, W1 = 2, W2 = 2, U_W1 = 3, U_W2 = 3)))

ggdag(dag, use_labels = "label") +
  theme_dag()


```

## TMLE Estimation

Use the `tmle` package to estimate a model for the effect of blood pressure medication on the probability of mortality. Do the following:

\begin{enumerate}
    \item Use the same SuperLearner library you defined earlier
    \item Use the same outcome model and propensity score model that you specified in the DAG above. If in your DAG you concluded that it is not possible to make a causal inference from this dataset, specify a simpler model and note your assumptions for this step.
    \item Report the average treatment effect and any other relevant statistics
\end{enumerate}

Some convenience assumptions are necessary for identification. We would need to assume that there are no unmeasured common causes $U$ between $A$ and $Y$ (e.g, $U_A \perp U_Y$) and $U_A \perp U_W$ __or__ $U_Y \perp U_W$ where $W$ represents each of our covariates). Although it is possible that unobserved demographic factors and unobserved health factors may have a causal relationship, we have chosen to maintain the independence assumption between $U_W1\perp U_W2$ so that we can proceed with the analysis.

```{r}
Y <- heart_disease$mortality
A <- heart_disease$blood_pressure_medication
W <- heart_disease %>% select(age, simplified_race, college_educ, income_thousands, bmi, blood_pressure, chol)

tmle_fit <-
  tmle::tmle(Y = Y, 
           A = A, 
           W = W, 
           Q.SL.library = SL.libraries, 
           g.SL.library = SL.libraries) 

tmle_fit
```

The average treatment effect of blood pressure medication is -0.355. The average treatment effect on the treated is -0.318. Both estimates are statistically significant at the $p<0.01$ level.

## Discussion Questions

\begin{enumerate}
    \item What is a "double robust" estimator? Why does it provide a guarantee of consistency if either the outcome model or propensity score model is correctly specified? Or in other words, why does mispecifying one of the models not break the analysis? \textbf{Hint}: When answering this question, think about how your introductory statistics courses emphasized using theory to determine the correct outcome model, and in this course how we explored the benefits of matching.
    A doubly robust estimator is consistent (and asymptotically unbiased) if either the outcome model $E_0(Y|A,W)$ or the propensity score $g_0(A|W)$ are correctly specified. Mispecifying one of these models does not break the analysis because of the law of large numbers. When $n$ is large, the sample average estimates the population average, making bias a second order term. If one of the bias terms in either model equals zero, taking the product of the two bias terms will also equal zero. Further, TMLE is an asymptotically (as $n \to \infty$) *efficient* estimator when both the outcome and exposure mechanisms are consistently estimated.
\end{enumerate}

# LTMLE Estimation

Now imagine that everything you measured up until now was in "time period 1". Some people either choose not to or otherwise lack access to medication in that time period, but do start taking the medication in time period 2. Imagine we measure covariates like BMI, blood pressure, and cholesterol at that time for everyone in the study (indicated by a "_2" after the covariate name). 

## Causal Diagram

Update your causal diagram to incorporate this new information. \textbf{Note}: If your groups divides up sections and someone is working on LTMLE separately from TMLE then just draw a causal diagram even if it does not match the one you specified above.

\textbf{Hint}: Check out slide 27 from Maya's lecture, or slides 15-17 from Dave's second slide deck in week 8 on matching.

\textbf{Hint}: Keep in mind that any of the variables that end in "\_2" are likely affected by both the previous covariates and the first treatment when drawing your DAG.

```{r}
# DAG for TMLE

# dag with combined demographic factors
dag <- dagify(Y1 ~ A0,
              Y1 ~ A1,
              Y1 ~ W_0,
              Y1 ~ W_1,
              A0 ~ W_0,
              A0 ~ W_1,
              A1 ~ W_0,
              A1 ~ W_1,
              W_1 ~ W_0,
              A1 ~ A0,
       exposure = c("A0", "A1"),
       outcome = "Y1",
      labels = c("Y1" = "mortality, t=1", "A0" = "bp meds, t=0", "A1" = "bp meds, t=1", "W_0" = "demog & health, t=0", "W_1" = "demog & health, t=1"),coords = list(x = c(A0 = -1.5, A1 = 2.5, Y1 = 3.25, W_0 = -1, W_1 = 2.5), y = c(A0 = -2, A1 = -2, Y1 = 3, W_0 = 1, W_1 = 1)))

ggdag(dag, use_labels = "label") +
  theme_dag()

```

## LTMLE Estimation

Use the `ltmle` package for this section. First fit a "naive model" that \textbf{does not} control for the time-dependent confounding. Then run a LTMLE model that does control for any time dependent confounding. Follow the same steps as in the TMLE section. Do you see a difference between the two estimates?

```{r}
## Naive Model (no time-dependent confounding) estimate

## LTMLE estimate
```

## Discussion Questions

\begin{enumerate}
    \item What sorts of time-dependent confounding should we be especially worried about? For instance, would we be concerned about a running variable for age the same way we might be concerned about blood pressure measured at two different times?
\end{enumerate}

